HyperLogLog: Exponentially Bad in Adversarial Settings

### 2021-04-24 docs/HLL_exp_bad/main.tex
new subsection to make impact on some system more clear
mods to proof

Both~\cite{cardestprivacy,hllvuln} propose mitigations using a salted HLL sketch, either as a replacement for (but at the cost of losing mergeability) or in addition to (inducing a memory overhead) an unsalted HLL. We show in Section~\ref{sec:formal} that when a salt is used appropriately (i.e.\ as the secret key for a pseudo-random function that is used in place of hash function $h$), security against adversarial inputs can be achieved even in our strongest adversarial setting. In addition, our simulation-based security definition, in which the simulator in the ideal world has no access to the inserted values, automatically guarantees confidentiality of the values inserted into the HLL sketch, even in the strongest attack model considered in~\cite{cardestprivacy}. So, while our main objective is to protect against adversarial inputs, we also get privacy for free when using a keyed PRF in place of the HLL hash function.

@@

\subsection{Relating the Experimental Results to Applications}\label{sec:exp-apps}

So far, we have treated attacks in an abstract way, working with an off-the-shelf implementation of HLL. We now illustrate what effect such attacks might have on applications relying on HLL for cardinality estimation. In an effort to be as specific a possible, we analyse in detail a recent paper~\cite{flexswitch} from the networking systems community.

The authors of~\cite{flexswitch} propose to use HLL for approximating traffic flow statistics in datacenter network switches. In the main example considered in~\cite[Section 3.1]{flexswitch}, each flow is identified by a 4-tuple consisting of source IP address, destination IP address, source port, and destination port. Various parameters for the HLL sketch to be used are then explored in~\cite{flexswitch}, trading memory consumed for accuracy of estimation of the number of concurrent flows. The paper also proposes to use the low range correction method of Fig.~\ref{fig:hll}. At no point does the discussion in~\cite{flexswitch} consider adversarial input (it does mention worst-case performance but only in passing, and no indication is given of how to handle this). It is clear that, by modifying the 16-bit source port (for example), an adversary could fool the switch into undercounting the flows it is responsible for producing. Depending on how the flow count is used, this could lead to undetected DoS conditions, mis-provisioning of network capacity, or under-billing of the adversary for its traffic.

More concretely,~\cite[Section 4.2]{flexswitch} proposes using HLL for detecting TCP port scans by counting observed destination port numbers in TCP packets with set SYN flags and reporting a scan event if the estimate exceeds a
set threshold. Using our attack techniques, an attacker could complete a partial port scan of roughly half of all $2^{16}$ ports whilst keeping the HLL estimate below $0.693m$ where $m$ is the number of buckets used in HLL. While~\cite{flexswitch} is not specific about the value of $m$ actually used, the authors' focus on using low memory indicates that $m$ could be $1024$.\footnote{The paper mentions using 1 byte counters in the buckets and states that using more than 1KB of memory ``only improves average error rates marginally''; we infer a choice of $m=1024$ from this.} It is not stated in~\cite{flexswitch} what threshold needs to be crossed to trigger an alarm, but we consider it plausible that it would be above $0.693 \cdot 1024 = 710$.

Similarly,~\cite[Section 4.2]{flexswitch} proposes using HLL for detecting NTP amplification attacks by counting distinct source IP addresses arriving on UDP port 123 over a small time window. No further details are given, but it is clear that an attacker with even limited flexibility in spoofing source IP addresses (e.g.\ within a private range used in the datacenter) could easily evade detection whilst delivering a large number of NTP requests.

In a more advanced application,~\cite[Section 4.3]{flexswitch} proposes to use HLL as a component in a novel version of the RCP congestion control algorithm called RCP-Approx. The idea of RCP-Approx is to replace exact counting with approximate counting to save on consumed memory in switches. Again HLL is used for estimating numbers of traffic flows. Obviously, our attack techniques could be used (in scenario S2) to cause RCP-Approx to underestimate the number of traffic flows, leading to non-detection of congestion and so enabling DoS attacks to go undetected.

@@

In this section we present a formal security analysis of HLL, focussing on whether we can augment the original HLL design with cryptographic mechanisms and thereby provably prevent attacks.

@@

\subsection{Formal Security Analysis for S2, S3, S4}

For scenarios S2, S3 and S4, the picture is much more positive. In the remainder of this section, we introduce a formal attack model covering all three scenarios. This is done by allowing an adversary ${\mathcal{A}}$ to interact with an HLL via various oracles that capture its capabilities in each of the three scenarios. In the security model, ${\mathcal{A}}$ interacts in one of two worlds. In the real world, ${\mathcal{A}}$ interacts with a real HLL, operating as in Figure~\ref{fig:hll}. In the ideal world, ${\mathcal{A}}$ interacts with an ideal HLL in which the operation $y: = h(x)$ is replaced by random sampling; that is, by setting $y \leftarrow_{\$} \{0,1\}^{n+\ell}$.

\todo{update! may need to formally define algs of an HLL as well as oracles. Since Sim will need to keep a shadow copy for itself.}

We then show that, if $h$ is replaced by a keyed Pseudo Random Function (PRF) $F$ whose key is not available to the adversary, then any adversary ${\mathcal{A}}$ that can distinguish between the real and random worlds with oracles as in S4 can be used to build an adversary ${{\mathcal{B}}}$ which breaks the PRF security of $F$. Under the assumption that $F$ is a good PRF, no such efficient adversary ${{\mathcal{B}}}$ can exist, and we may conclude that we can safely switch from the real world to the ideal world without incurring any significant change to the behaviour of ${\mathcal{A}}$.

However, it is clear that any adversary in the ideal world is reduced to operating in the setting used in the average case analysis of HLL (wherein every insertion query results in a random value $y$ being used to update the HLL sketch). Thus, ${\mathcal{A}}$ can have no advantage when it comes to influencing the HLL cardinality estimate. It follows that HLL, when instantiated with PRF $F$, is secure against attacks in S4 (and therefore in the weaker attacks models S2 and S3 as well). We will conclude by discussing how $F$ can be instantiated.

\todo{explain the above better with simulators + conf for free.}

We note that our formal security analysis could be conducted using the framework recently laid out in~\cite{CCS:ClaPatShrPS19}. However, we prefer to give a compact and self-contained treatment here.

@@

\todo{Why UHFs don't work in general -- key recovery is possible in S4; partial info about processed values and therefore keys  leaks from Card Est in S2, though harder to make an attack from this. }

\todo{Upgrade to use simulator S as a typical simulation-based security notion should; reference to Lindell's ``How to simulate it'' paper? Order of quantifiers; input to simulator: nothing, not even number of queries, assuming adversary never repeats an insert query.}

We consider adversary ${\mathcal{A}}$ to be an algorithm running against a challenger ${{\mathcal{C}}}$ in a security game. The game begins with ${{\mathcal{C}}}$ selecting a bit $b$ uniformly at random. Adversary ${\mathcal{A}}$ interacts with up to four oracles which are handled by ${{\mathcal{C}}}$; different combinations of oracles define scenarios S2, S3, S4. After making some queries, ${\mathcal{A}}$ returns a bit $b'$ estimating whether it is interacting with a real HLL instantiation ($b'=0$) or an ideal one ($b'=1$). For $i=1, 2, 3$, we define ${\mathcal{A}}$'s S$i$-Real-or-Ideal advantage against HLL to be:
\[
\mbox{Adv}^{\mbox{\tiny S$i$-RoI}}_{\mbox{\tiny HLL}}({\mathcal{A}}) := \lvert 2\Pr[b' = b] - 1 \rvert \, .
\]
%Using a standard transformation, this probability can be rewritten as $\lvert \Pr[b' = 1| b = 1] - \Pr[b'=0|b=1] \rvert$. 
The four oracles that  ${\mathcal{A}}$ can interact with are as follows:
\begin{itemize}
\item \texttt{init}: this oracle performs initialisation of the HLL sketch (and selection of the PRF key when $h$ is replaced by a PRF $F$); this oracle is available in all of S2, S3, S4, but can only be called once as ${\mathcal{A}}$'s first oracle query.
\item \texttt{insert($\cdot$)}: on input $x \in {{\mathcal{D}}}$, the oracle inserts value $x$ into the HLL sketch. This oracle is available in all of S2, S3, S4. Without loss of generality (since HLL and variants of it we consider here have a deterministic insertion procedure), we assume all queries $x$ made by ${\mathcal{A}}$ to \texttt{insert($\cdot$)} are distinct. We let $q$ denote the number of insertion queries made by ${\mathcal{A}}$ during the course of its attack.
\item \texttt{read}: this oracle returns the HLL cardinality estimate. We allow multiple queries to this oracle and make it available to ${\mathcal{A}}$ in S3 and S4, but not S2.
\item \texttt{corrupt}: this oracle returns the vector $(M[j])_{j=1}^{m}$ of HLL bucket contents (but not any secret key $K$). This oracle is available only in S4. We allow the adversary to query \texttt{corrupt} as often as it wishes during its attack. Our description of S4 in Section~\ref{sec:attacks} limited the adversary to at most one such query; we allow multiple queries here because doing so does not increase the difficulty of achieving security.  
\end{itemize}

\subsection{Security Analysis of HLL with a PRF}

We now replace the hash function $h$ acting on ${{\mathcal{D}}}$ from the HLL algorithm with a keyed function $F: \{0,1\}^k \times {{\mathcal{D}}} \rightarrow  \{0,1\}^{n+\ell}$. Here the first input to $F$ is a $k$-bit string, the key, denoted $K$. We assume that \texttt{init} in the HLL security model now initialises key $K$ by selecting it uniformly at random from $\{0,1\}^k$.

Now we are ready to state:

Informally, if $F$ is a good PRF, then $\mbox{Adv}^{\mbox{\tiny PRF}}_{F}({{\mathcal{B}}})$ should be small for any adversary ${{\mathcal{B}}}$ running in a reasonable amount of time; hence the advantage of any S4 adversary  ${\mathcal{A}}$ against HLL using PRF $F$ is also small. Note that our proof still applies for the improved algorithm of~\cite{hllnew} that is used in Redis. The only changes to the proof are to use the Redis approach for computing responses to \texttt{read} queries in place of the classical HLL approach.

\begin{proof}
PRF adversary ${{\mathcal{B}}}$ against $F$ is built as follows. Let $d$ denote the hidden bit determining whether ${{\mathcal{B}}}$ receives real responses of the form $F(K,x)$ where $K$ is a uniformly random key ($d=0$) or responses from a truly random function ($d=1$); ${{\mathcal{B}}}$ runs S4 adversary ${\mathcal{A}}$ and responds to each of its queries as follows:
\begin{itemize}
\item \texttt{init}: here ${{\mathcal{B}}}$ initialises an HLL sketch in accordance with Figure~\ref{fig:hll}. Thus it initialises an array $M[j]$ in which each entry is set to 0. 
\item \texttt{insert($\cdot$)}: on input $x \in {{\mathcal{D}}}$,  ${{\mathcal{B}}}$ makes a query on $x$ to its real-or-random oracle and receives value $y$; ${{\mathcal{B}}}$ uses $y$ to update the HLL bucket $j$ generated via $j=1 + <y_1y_2...y_{n}>_2$ as in the regular HLL algorithm (shown in Figure~\ref{fig:hll}).
\item \texttt{read}: ${{\mathcal{B}}}$ uses its knowledge of the buckets' contents to compute and return the HLL cardinality estimate in accordance with Figure~\ref{fig:hll}. 
\item \texttt{corrupt}: ${{\mathcal{B}}}$ returns the vector $(M[j])_{j=1}^{m}$ of HLL bucket contents. 
\end{itemize}
When ${\mathcal{A}}$ outputs a bit $b'$, ${{\mathcal{B}}}$ outputs its own bit $d'=b'$. Note that when $d=0$, ${{\mathcal{B}}}$ plays against the real $F$ with a random key $K$ and provides a simulation for ${\mathcal{A}}$ that perfectly instantiates the real world. On the other hand, when $d=1$, ${{\mathcal{B}}}$ plays against a truly random function and provides a simulation for ${\mathcal{A}}$ that perfectly instantiates the ideal world. The equality of advantages in the statement of the theorem now follows immediately. The remainder of the proof follows on counting queries and noting that the HLL algorithm for computing cardinality is efficient, so ${{\mathcal{B}}}$'s running time due to handling \texttt{read} queries is not increased significantly compared to that of ${\mathcal{A}}$.
\end{proof}

We conclude by reiterating our central point: we have shown that no efficient S4 adversary ${\mathcal{A}}$ can tell that it is playing against HLL implemented using truly random values $y$ instead of with values obtained from a PRF, except with a small advantage. This means that ${\mathcal{A}}$ is effectively reduced to attacking HLL in the exact same setting used in the average case analysis of HLL from~\cite{hll}, wherein it is proved that the HLL cardinality estimate endures relative errors of the form $1.04/\sqrt{m}$. 
It does not matter that ${\mathcal{A}}$ can adaptively choose its inputs $x$ in \texttt{insert($\cdot$)} queries, nor that it can access \texttt{read} and \texttt{corrupt} queries. For each \texttt{insert($\cdot$)} query, a fresh value $x$ leads to a new random $y$ being generated and used to update a bucket, as in the average case analysis of HLL. The \texttt{read} and \texttt{corrupt} queries do leak information about $F$'s outputs to the adversary, but such outputs are indistinguishable from random under the assumption that $F$ is a PRF.

\todo{revisit the above in view of introduction of simulator, plus the confidentiality aspect for free. NB corruption query does not give PRF key.}

@@

The above construction requires a PRF $F: \{0,1\}^k \times {{\mathcal{D}}} \rightarrow  \{0,1\}^{n+\ell}$. Here, the set ${{\mathcal{D}}}$ represents the space of possible inputs to HLL which can be modelled by a set of bit-strings of some maximum length. This means that $F$ needs to be Variable Input Length PRF (VIL-PRF).

@@

We have given a formal analysis of using a PRF as a countermeasure to our attacks. The use of a keyed function sacrifices HLL mergeability when the keys in the different HLLs are independently generated. It is an open problem to find methods for securing HLL sketches against adversarial inputs whilst retaining some form of mergeability. This may also require the introduction of entirely new data structures that trade increased storage for mergeability and improved security. 
%It is also an open problem to close the performance gap that exists between fast but non-cryptographic hashes used in HLL implementations today and the slower but cryptographically strong VIL-PRF that is needed to obtain security.

>>>>

Both~\cite{cardestprivacy,hllvuln} propose mitigations using a salted HLL sketch, either as a replacement for (but at the cost of losing mergeability) or in addition to (inducing a memory overhead) an unsalted HLL. We show in Section~\ref{sec:formal} that when a salt is used appropriately (i.e.\ as the secret key for a pseudo-random function that is used in place of hash function $h$), security against adversarial inputs can be achieved even in our strongest adversarial setting. 
%In addition, our simulation-based security definition, in which the simulator in the ideal world has no access to the inserted values, automatically guarantees confidentiality of the values inserted into the HLL sketch, even in the strongest attack model considered in~\cite{cardestprivacy}. So, while our main objective is to protect against adversarial inputs, we also get privacy for free when using a keyed PRF in place of the HLL hash function.

@@

\subsection{Relating the Attacks to Applications}\label{sec:exp-apps}

So far, we have treated attacks in an abstract way, working with an off-the-shelf implementation of HLL. We now illustrate what effect such attacks might have on applications relying on HLL for cardinality estimation. In an effort to be more concrete, we analyse in detail a recent paper~\cite{flexswitch} from the networking systems community.

The authors of~\cite{flexswitch} propose to use HLL for recording traffic flow statistics in datacenter network switches. The idea is to use HLL so as to reduce the amount of memory needed in the switches. In the main example considered in~\cite[Section 3.1]{flexswitch}, each flow is identified by a 4-tuple consisting of source IP address, destination IP address, source port, and destination port. Various parameters for the HLL sketch to be used are then explored in~\cite{flexswitch}, trading memory consumed for accuracy of estimation of the number of concurrent flows. The paper also proposes to use the low range correction method of Fig.~\ref{fig:hll}. At no point does the discussion in~\cite{flexswitch} consider adversarial input (it does mention worst-case performance but only in passing, and no indication is given of how to handle this). It is clear that, by modifying the 16-bit source port (for example), an adversary could fool the switch into undercounting the flows the adversary is responsible for producing. Depending on how the flow count is used, this could lead to undetected DoS conditions, mis-provisioning of network capacity, or under-billing of the adversary for its traffic.

Going further, \cite[Section 4.2]{flexswitch} proposes using HLL for detecting TCP port scans by counting observed destination port numbers in TCP packets with set SYN flags and reporting a scan event if the estimate exceeds a
set threshold. Using our attack techniques, an attacker could complete a partial port scan of roughly half of all $2^{16}$ ports whilst keeping the HLL estimate below $0.693m$ where $m$ is the number of buckets used in HLL. While~\cite{flexswitch} is not specific about the value of $m$ actually used, the authors' focus on reducing memory usage indicates that $m$ would not be larger than $1024$.\footnote{The paper mentions using 1 byte counters in the buckets and states that using more than 1KB of memory ``only improves average error rates marginally''; we infer a maximum value of $m=1024$ from this.} It is not stated in~\cite{flexswitch} what threshold needs to be crossed to trigger an alarm, but we consider it plausible that it would be above $0.693 \cdot 1024 = 710$.

Similarly,~\cite[Section 4.2]{flexswitch} proposes using HLL for detecting NTP amplification attacks by counting distinct source IP addresses arriving on UDP port 123 over a small time window. No further details are given, but it is clear that an attacker with even limited flexibility in spoofing source IP addresses (e.g.\ within a private range used in the datacenter) could easily evade detection whilst delivering a large number of NTP requests, thus re-enabling the original attack.

In a more advanced application,~\cite[Section 4.3]{flexswitch} proposes to use HLL as a component in a novel, low-memory version of the RCP congestion control algorithm called RCP-Approx. Again HLL is used for estimating numbers of traffic flows. Obviously, our attack techniques could be used (in scenario S2) to cause RCP-Approx to underestimate the number of traffic flows, leading to non-detection of congestion and so enabling DoS attacks to go undetected.

@@

In this section we present a formal security analysis of HLL, focussing on how to augment the original HLL design with cryptographic mechanisms and thereby provably prevent attacks.

@@

\subsection{Intuition for Formal Security Analysis for S2, S3, S4}

For scenarios S2, S3 and S4, the picture is much more positive. In the remainder of this section, we introduce a formal security model covering all three scenarios. We work with a simulation-based framework. This is in contrast to the game-based approach introduced for probabilistic data structures in~\cite{CCS:ClaPatShrPS19}. We discuss the strengths of our alternative approach in more detail below.

We first begin by explaining the high-level idea of our approach. We allow an adversary ${\mathcal{A}}$ to interact with an HLL via various oracles that capture its capabilities in each of the three scenarios S2, S3 and S4: it may have an \texttt{insert($\cdot$)} oracle allowing it to add values to the HLL, a \texttt{read} oracle that returns the HLL cardinality estimate, and a \texttt{corrupt} that returns all the HLL bucket contents (but not any secret key $K$); in S4 it has all three oracles. In addition, ${\mathcal{A}}$ always has access to an \texttt{init($\cdot$)} oracle by which it receives the HLL parameters $n$, $\ell$, $\alpha_n$ needed to initialise the HLL sketch.

In the security model, ${\mathcal{A}}$ interacts in one of two worlds. In the real world, ${\mathcal{A}}$ interacts with a real HLL, operating as in Figure~\ref{fig:hll} but with the hash function $h$ replaced by a keyed function $F: \{0,1\}^k \times {{\mathcal{D}}} \rightarrow  \{0,1\}^{n+\ell}$ whose key is not available to the adversary. In the ideal world, ${\mathcal{A}}$ interacts with a simulator ${\mathcal{S}}$ which ignores the input for ${\mathcal{A}}$'s \texttt{insert($\cdot$)} queries but nevertheless provides a ``good'' simulation of the oracle responses to ${\mathcal{A}}$. Importantly, ${\mathcal{S}}$ will handle all of ${\mathcal{A}}$'s requests to insert an item $x$ using \emph{random sampling}, that is, by setting $y \leftarrow_{\$} \{0,1\}^{n+\ell}$ instead of $y: = F_K(x)$.

We will show that when $h$ is replaced by the keyed function $F$	 then any adversary ${\mathcal{A}}$ that can distinguish between the real and ideal worlds can be used to build an adversary ${{\mathcal{B}}}$ which breaks the PRF security of $F$. Under the assumption that $F$ is a good PRF, no such efficient adversary ${{\mathcal{B}}}$ can exist, and we may conclude that (up to PRF security) ${\mathcal{A}}$ cannot tell that it is operating in the ideal world instead of the real world.

But in the ideal world, ${\mathcal{A}}$ interacts with a simulator that uses random sampling for the $y$ values. This effectively reduces ${\mathcal{A}}$ to operating in the setting where the average case analysis of HLL applies, as per~\cite{hll}. Here, ${\mathcal{A}}$ has no advantage when it comes to influencing the HLL cardinality estimate. It follows that HLL, when instantiated with PRF $F$, is secure against attacks in S4 (and therefore in the weaker attacks models S2 and S3 as well). 
%Recall also that the simulator ${\mathcal{S}}$ ignores the values $x$ being inserted by ${\mathcal{A}}$. This enables us to get certain confidentiality properties for free when HLL is implemented using a PRF in place of $h$. For example, we can argue that the bucket contents (obtained via a \texttt{corrupt} query) reveal nothing to ${\mathcal{A}}$ about what values $x$ were inserted into the HLL.

We will conclude our analysis by discussing how $F$ can be instantiated.

We consider adversary ${\mathcal{A}}$ to be an algorithm running in a security game. The game begins with the selection of a bit $b$ uniformly at random. Adversary ${\mathcal{A}}$ interacts with up to four oracles which are handled by the game; different combinations of oracles define scenarios S2, S3, S4. After making some queries, ${\mathcal{A}}$ returns a bit $b'$ estimating whether it is interacting with a real HLL instantiation ($b'=0$) or an ideal one ($b'=1$). For $i=1, 2, 3$, we define ${\mathcal{A}}$'s S$i$-Real-or-Ideal advantage against HLL to be:
\[
\mbox{Adv}^{\mbox{\tiny S$i$-RoI}}_{\mbox{\tiny HLL}}({\mathcal{A}}) := \lvert 2\Pr[b' = b] - 1 \rvert \, .
\]
%Using a standard transformation, this probability can be rewritten as $\lvert \Pr[b' = 1| b = 1] - \Pr[b'=0|b=1] \rvert$. 
The four oracles that  ${\mathcal{A}}$ can interact with behave as follows in the real world:
\begin{itemize}
\item \texttt{init($\cdot$)}: this oracle receives as input the HLL parameters ($n$, $\ell$, $\alpha_n$) and performs initialisation of the HLL sketch. It also selects a key $K \leftarrow_{\$} \{0,1\}^k$ for keyed function $F: \{0,1\}^k \times {{\mathcal{D}}} \rightarrow  \{0,1\}^{n+\ell}$. This oracle is available in all of S2, S3, S4, but can only be called once as ${\mathcal{A}}$'s first oracle query.
\item \texttt{insert($\cdot$)}: on input $x \in {{\mathcal{D}}}$, the oracle sets $y = F_K(x)$ and follows the procedure in Figure~\ref{fig:hll} to insert value $x$ into the HLL sketch (that is, it uses the first $m$ bits of $y$ to determine a bucket index $j$ and the next $\ell$ bits of $y$ to determine a bit-string $w$; then $\rho(w)$ is used to update the bucket at index $j$ via $M[j]:= \max(M[j],\rho(w))$). This oracle is available in all of S2, S3, S4. Without loss of generality (since HLL and variants of it we consider here have a deterministic insertion procedure), we assume all queries $x$ made by ${\mathcal{A}}$ to \texttt{insert($\cdot$)} are distinct. We let $q$ denote the number of insertion queries made by ${\mathcal{A}}$ during the course of its attack.
\item \texttt{read}: this oracle returns the HLL cardinality estimate as per Figure~\ref{fig:hll}. We allow multiple queries to this oracle and make it available to ${\mathcal{A}}$ in S3 and S4, but not S2.
\item \texttt{corrupt}: this oracle returns the vector $(M[j])_{j=1}^{m}$ of HLL bucket contents (but not any secret key $K$). This oracle is available only in S4. We allow the adversary to query \texttt{corrupt} as often as it wishes during its attack. Our description of S4 in Section~\ref{sec:attacks} limited the adversary to at most one such query; we allow multiple queries here because doing so does not increase the difficulty of achieving security.  
\end{itemize}

In many simulation-based security proofs, the simulator ${\mathcal{S}}$ depends on ${\mathcal{A}}$, and security is quantified via a statement of the form ``For all adversaries ${\mathcal{A}}$, there exists a simulator ${\mathcal{S}}$ such that $\ldots$'' with ${\mathcal{S}}$ being described in the proof. In our analysis we are able to provide a universal simulator, enabling us to make a stronger statement of the form ``There exists a simulator ${\mathcal{S}}$ such that, for all adversaries ${\mathcal{A}}\ldots$''. We describe this simulator now, in terms of how it handles  ${\mathcal{A}}$'s oracle queries in the ideal world:

\begin{itemize}
\item \texttt{init($\cdot$)}: on input $n$, $\ell$, $\alpha_m$, ${\mathcal{S}}$ initialises an HLL sketch as in Figure~\ref{fig:hll}; that is, it creates an array of $m=2^n$ buckets $M[j]$ and initialises them all to 0.
\item \texttt{insert($\cdot$)}: on input $x \in {{\mathcal{D}}}$,  ${\mathcal{S}}$ ignores $x$, sets $y \leftarrow_{\$} \{0,1\}^{n+\ell}$, and proceeds to update the buckets using this value of $y$ as in the original HLL algorithm in Figure~\ref{fig:hll}.
\item \texttt{read}:  ${\mathcal{S}}$ returns the HLL cardinality estimate computed as per Figure~\ref{fig:hll}.
\item \texttt{corrupt}: ${\mathcal{S}}$ returns the current vector $(M[j])_{j=1}^{m}$ of HLL bucket contents that it has stored.
\end{itemize}

Now we are ready to state our main result:

%Informally, if $F$ is a good PRF, then $\mbox{Adv}^{\mbox{\tiny PRF}}_{F}({{\mathcal{B}}})$ should be small for any adversary ${{\mathcal{B}}}$ running in a reasonable amount of time; hence the advantage of any S4 adversary  ${\mathcal{A}}$ against HLL using PRF $F$ is also small. Note that our proof still applies for the improved algorithm of~\cite{hllnew} that is used in Redis. The only changes to the proof are to use the Redis approach for computing responses to \texttt{read} queries in place of the classical HLL approach.

\begin{proof}
PRF adversary ${{\mathcal{B}}}$ against $F$ is built as follows. Let $d$ denote the hidden bit determining whether ${{\mathcal{B}}}$ receives real responses of the form $F_K(x)$ where $K$ is a uniformly random key ($d=0$) or responses from a truly random function ($d=1$); ${{\mathcal{B}}}$ runs S4 adversary ${\mathcal{A}}$ and responds to each of its queries as follows:
\begin{itemize}
\item \texttt{init}: here ${{\mathcal{B}}}$ initialises an HLL sketch in accordance with Figure~\ref{fig:hll}. Thus it initialises an array $M[j]$ in which each entry is set to 0. 
\item \texttt{insert($\cdot$)}: on input $x \in {{\mathcal{D}}}$,  ${{\mathcal{B}}}$ makes a query on $x$ to its real-or-random oracle and receives value $y$; ${{\mathcal{B}}}$ uses $y$ to update the HLL bucket $j$ generated via $j=1 + <y_1y_2...y_{n}>_2$ as in the regular HLL algorithm (shown in Figure~\ref{fig:hll}).
\item \texttt{read}: ${{\mathcal{B}}}$ uses its knowledge of the buckets' contents to compute and return the HLL cardinality estimate in accordance with Figure~\ref{fig:hll}. 
\item \texttt{corrupt}: ${{\mathcal{B}}}$ returns the vector $(M[j])_{j=1}^{m}$ of HLL bucket contents. 
\end{itemize}
When ${\mathcal{A}}$ outputs a bit $b'$, ${{\mathcal{B}}}$ outputs its own bit $d'=b'$. Note that when $d=0$, ${{\mathcal{B}}}$ plays against the real $F$ with a random key $K$ and provides a simulation for ${\mathcal{A}}$ that perfectly instantiates the real world. On the other hand, when $d=1$, ${{\mathcal{B}}}$ plays against a truly random function and provides a simulation for ${\mathcal{A}}$ that perfectly instantiates the ideal world with simulator $\mathcal{S}$ (to see this, note that sampling the outputs $y$ of a truly random function at distinct inputs $x$ is identical to simply sampling $y$ uniformly at random). The equality of advantages in the statement of the theorem now follows immediately. The remainder of the proof follows on counting queries and noting that the HLL algorithm for computing cardinality is efficient, so ${{\mathcal{B}}}$'s running time due to handling \texttt{read} queries is not increased significantly compared to that of ${\mathcal{A}}$.
\end{proof}

We reiterate our central point: the theorem shows that no efficient S4 adversary ${\mathcal{A}}$ can tell that it is playing against our simulator ${\mathcal{S}}$ using truly random values $y$ instead of with values obtained from a PRF, except with a small advantage. This means that ${\mathcal{A}}$ is effectively reduced to attacking HLL in the exact same setting used in the average case analysis of HLL from~\cite{hll}, wherein it is proved that the HLL cardinality estimate endures relative errors of the form $1.04/\sqrt{m}$. From this we can conclude that no efficient S4 adversary ${\mathcal{A}}$ can manipulate the HLL cardinality estimate to its benefit.

\subsection{On Confidentiality}
We also recall that our ${\mathcal{S}}$ ignores ${\mathcal{A}}$'s inputs $x$ in handling \texttt{insert($\cdot$)} queries. In traditional simulation-based security notions, this would imply that the ``leakage'' given to the simulator does not include the input values $x$, and one would then conclude by analogy to, say, semantic security for encryption that the HLL implemented using a PRF offers confidentiality for the inputs $x$. This is not quite correct here, since ${\mathcal{A}}$ chooses all the inputs $x$ and so ``knows'' them already. To address confidentiality, one would need instead to consider $x$ that are drawn from some distribution on $\mathcal{D}$ or consider two-stage adversaries that ``forget'' the $x$ values chosen in its first stage by not passing state to its second stage.

Even here, however, an HLL can leak information about values $x$ that were previously inserted. Consider the following attack from~\cite{cardestprivacy}: to test whether a value $x$ was previously inserted into the HLL, make the sequence of queries: \texttt{read}, \texttt{insert($x$)}, \texttt{read}. If the two cardinality estimates are different, one can conclude definitively that $x$ was \emph{not} previously inserted into the HLL; if they are the same, one can estimate (with some error) that $x$ was previously inserted. This attack still works even if HLL is implemented using a PRF and is unavoidable in view of the oracle queries (and in real settings, the API) available to the adversary. So such privacy leaks are an inherent feature of HLL if a suitably rich API is provided. Note that this attack does not contradict our main result, which concerns prevention of cardinality manipulation rather than confidentiality.

More positively, what we can say is that our simulation-based security definition does imply confidentiality of inputs in the  case of a model where an honest party makes a sequence of insertions and where ${\mathcal{A}}$ is then limited to \texttt{read} and \texttt{corrupt} queries. \todo[inline]{explain why.}

We leave the further exploration of how to adapt our simulation-based security definitions to capture confidentiality notions to future work.

%It does not matter that ${\mathcal{A}}$ can adaptively choose its inputs $x$ in \texttt{insert($\cdot$)} queries, nor that it can access \texttt{read} and \texttt{corrupt} queries. For each \texttt{insert($\cdot$)} query, a fresh value $x$ leads to a new random $y$ being generated and used to update a bucket, as in the average case analysis of HLL. The \texttt{read} and \texttt{corrupt} queries do leak information about $F$'s outputs to the adversary, but such outputs are indistinguishable from random under the assumption that $F$ is a PRF.

@@

\todo{Why UHFs don't work in general so why we need a PRF -- key recovery is possible in S4; partial info about processed values and therefore keys  leaks from Card Est in S2, though harder to make an attack from this. }

@@

Our HLL construction requires a PRF $F: \{0,1\}^k \times {{\mathcal{D}}} \rightarrow  \{0,1\}^{n+\ell}$. Here, the set ${{\mathcal{D}}}$ represents the space of possible inputs to HLL which can be modelled by a set of bit-strings of some maximum length. This means that $F$ needs to be Variable Input Length PRF (VIL-PRF).

@@

We have given a formal analysis of using a PRF as a countermeasure to our attacks. The use of a keyed function sacrifices HLL mergeability when the keys in the different HLLs are independently generated. It is an open problem to find methods for securing HLL sketches against adversarial inputs whilst retaining some form of mergeability. This may also require the introduction of entirely new data structures that trade increased storage for mergeability and improved security.

\todo{Sim based of indepednent interest, arguably more powerful than game-based notions with fixed winning predicates.}
%It is also an open problem to close the performance gap that exists between fast but non-cryptographic hashes used in HLL implementations today and the slower but cryptographically strong VIL-PRF that is needed to obtain security.
